---
title: "Analysis of calibrated model stable states"
output: html_document
date: '2023-01-17'
---

# Analysis of calibrated model stable states

Code used to investigate the stable state of calibrated models. All stable states are visualised in heatmaps. Additional compliance plots are created to assess the compliance between stable state and baseline activity profile.

### Import necessary libraries

```{r}
library(emba)
library(pheatmap)
library(gtools)
library(data.table)
library(ggplot2)
```

### Define relevant variable names

```{r}
pipeline_folder_name = readline(prompt="Enter the name of the pipeline folder: ")
baseline_activity_file = readline(prompt="Enter the name of the baseline activity profile: ")
```

### *import_baseline_activity_profile():* Import and process the baseline activity profile

```{r}
import_baseline_activity_profile = function(baseline_activity_profile){
  baseline_activity_data = read.table(baseline_activity_profile, header = TRUE)
  baseline_activity_data$Node = as.character(baseline_activity_data$Node)
  baseline_activity_data[nrow(baseline_activity_data)+1,] = c('Antisurvival',0)
  baseline_activity_data[nrow(baseline_activity_data)+1,] = c('Prosurvival',1)
  rownames(baseline_activity_data) = baseline_activity_data$Node
  baseline_activity_data = subset(baseline_activity_data, select = -c(Node))
  colnames(baseline_activity_data) = c('baseline_activity')
  return(baseline_activity_data)
}
```

### *single_simulation_df():* Import stable state of model ensemble and store as dataframe

```{r}
single_simulation_data_frame = function(simulation_name, pipeline_folder){
  
  path = paste0(getwd(),'/druglogics-synergy/',pipeline_folder,'/',simulation_name,'/models')
  
  ss = get_stable_state_from_models_dir(models.dir = path, all.ss = TRUE)
  ss = ss[,-ncol(ss)]
  
  return(ss)
}
```

### *single_simulation_heatmap(): Create stable state of model ensemble*

```{r}
single_simulation_heatmap = function(simulation_name, baseline_activity_profile, plot_title){
  
  path = paste0(getwd(),'/druglogics-synergy/',pipeline_folder,'/',simulation_name,'/models')
  
  sim_df = single_simulation_df(simulation_name, pipeline_folder)
  sim_df = t(single_df)
  
  baseline_activity_data = import_baseline_activity_profile(baseline_activity_profile)
  
  #create a heclustering of columns
  sim_heatmap = pheatmap(sim_df, fontsize = 16, fontsize_row = 8, fontsize_col = 6, show_rownames = TRUE, cluster_rows = TRUE, annotation_legend = TRUE, fontfamily = "serif", main = plot_title, annotation_row = baseline_activity_data, annotation_colors = list(baseline_activity = c('1' = "red2", 'NA' = "white", "0.5" = "white", '0' = "royalblue1")), annotation_names_row = FALSE)
  
  return(sim_heatmap)
}
```

### *single_simulation_compliance():* Calculate the compliance between the average model ensemble stable state and the baseline activity profile

```{r}
single_simulation_compliance = function(simulation_name, baseline_activity_profile, pipeline_folder){
  
  sim_df = single_simulation_df(simulation_name, pipeline_folder)
  baseline_activity_data = import_baseline_activity_profile(baseline_activity_profile)
  
  single_compliance = c()
  
  for(n in 1:(ncol(sim_df)-1)){
      node_df = sim_df[[n]]
      average = sum(node_df)/nrow(sim_df)
      baseline_value = as.integer(baseline_activity_data[colnames(sim_df[n]), ])
      if(is.na(baseline_value)){
        difference = 'NA'
      }
      else{
      difference = 1-(abs(baseline_value-average))
      single_compliance = append(difference,single_compliance)
      }
  }
  
  compliance = sum(single_compliance)/length(single_compliance)
  return(compliance)
}
```

### *across_simulation_data_frame():* Import stable state of model ensembles across a selection of simulations and store in data frame

```{r}
#Function used to generate a data frame includng all attractors reached by the model ensemble.
across_simulation_data_frame = function(mode, pipeline_folder){
  sim_path = paste0(getwd(),'/druglogics-synergy/',pipeline_folder)
  sim_names = mixedsort(dir(path = sim_path, pattern = paste0(mode,'_',pipeline_folder)))
  df_total = data.frame()
  for(s in sim_names){
    ss = get_stable_state_from_models_dir(models.dir = paste0(sim_path,'/',s,'/models'), all.ss = TRUE)
    for(n in 1:(ncol(ss)-1)){
      node_ss = ss[[n]]
      average = sum(node_ss)/nrow(ss)
      ss[[n]][1] = average
    }
    df_total = rbind(df_total,ss[1,])
  }
  df_total$model_name = c(paste0(mode,"_10"), paste0(mode,"_20"), paste0(mode,"_30"), paste0(mode,"_40"), paste0(mode,"_50"), paste0(mode,"_60"), paste0(mode,"_70"), paste0(mode,"_80"), paste0(mode,"_90"), paste0(mode,"_100"))
  row_names = df_total$model_name
  df_total = df_total[, -ncol(df_total)]
  rownames(df_total) = row_names
  df_total = t(df_total)
  return(df_total)
}
```

### *across_simulation_heatmap():* Visualise model ensemble stable state across a selection of simulations

```{r}
across_simultation_heatmap = function(baseline_activity_profile, mode, pipeline_folder){

  across_simulation_df = across_simulation_data_frame(mode, pipeline_folder)
  baseline_activity = import_baseline_activity_profile(baseline_activity_profile)

  output_heatmap = pheatmap(across_simulation_df, fontsize = 16, fontsize_row = 8, fontsize_col = 14, cluster_rows = TRUE, cluster_cols = FALSE, annotation_row = baseline_activity, annotation_colors = list(baseline_activity = c('1' = "red2", 'NA' = "white", '0.5' = "white", '0' = "royalblue1")), annotation_legend = TRUE, fontfamily = "serif", main = "Stable states of calibrated models", annotation_names_row = FALSE)
  return(output_heatmap)
}
```

### *across_simulation_compliance():* Calculate compliance with baseline activity profile for a selection of simulations

```{r}
across_simulation_compliance = function(baseline_activity_profile, mode, pipeline_folder){
  
  simulation_df = across_simulation_data_frame(mode, pipeline_folder)
  
  baseline_activity = import_baseline_activity_profile(baseline_activity_profile)
  baseline_activity$baseline_activity = as.double(baseline_activity$baseline_activity)
  
  nodes = rownames(simulation_df)
  
  simulation_compliance = data.frame(matrix(nrow = 1, ncol = 2))
  
  for(s in colnames(simulation_df)){
    compliance = data.frame(matrix(nrow = 1, ncol = 2))
    for (i in nodes){
      value = simulation_df[i,s]
      activity = baseline_activity[i,]
      if(is.na(activity)){
        difference = 'NA'
        info = c(i,difference)
      }
      else{
        difference = 1-(abs(activity-value))
        info = c(i,difference)
        compliance = rbind(compliance, info)
      }
    }
    colnames(compliance) = c("Node", "Compliance")
    compliance = compliance[-1,]
    compliance$Compliance = as.double(compliance$Compliance)
    compliance_average = round(sum(compliance$Compliance)/nrow(compliance), 3)
    sim_info = c(s,compliance_average)
    simulation_compliance = rbind(simulation_compliance, sim_info)
  }
  
  colnames(simulation_compliance) = c('Simulation','Compliance')
  simulation_compliance = simulation_compliance[-1,]
  simulation_compliance$size = c(10,20,30,40,50,60,70,80,90,100)
  simulation_compliance$Compliance = as.double(simulation_compliance$Compliance)
  
  return(simulation_compliance)
}
```

### *consistent_nodes(): Check for consistent nodes in stable states across simulations*

```{r}
#Check if the node activity is consistent across all models in the model ensemble
consistent_nodes = function(df){
  for (n in 1:nrow(df)){
    if(sum(df[n,]) == ncol(df)){
      conclusion = paste0(row.names(df)[n], ' is consistently active')
      print(conclusion)
    }
    else if(sum(df[n,]) == 0){
      conclusion = paste0(row.names(df)[n], ' is consistently inactive')
      print(conclusion)
    }
  }
}
```

### Stable state analysis of random model

```{r}
random_data_frame = single_simulation_data_frame("random_model_simulation", pipeline_folder_name)

random_model_heatmap = single_simulation_heatmap("random_model_simulation", baseline_activity_file, "Stable state of random model")

random_model_compliance = single_simulation_compliance("random_model_simulation", baseline_activity_file, pipeline_folder_name)
```

### Stable state analysis of simulations using DP-based calibration datasets

```{r}
DP_data_frame = across_simulation_data_frame("DP", pipeline_folder_name)
DP_heatmap = across_simultation_heatmap(baseline_activity_file, "DP", pipeline_folder_name)
```

```{r}
consistent_nodes(DP_data_frame)
```

```{r}
DP_compliance = across_simulation_compliance(baseline_activity_file, "DP", pipeline_folder_name)

DP_compliance_plot = ggplot(DP_compliance, aes(x=size, y=Compliance, label = Compliance))+
  geom_point()+
  geom_hline(yintercept = random_model_compliance, color = "red", linetype = "dashed")+
  geom_text(aes(x = 100, y = random_model_compliance - 0.05, label = "Random model"), size = 3, family='serif', color = "red")+
  theme(text = element_text(size = 12, family = "serif"))+
  geom_text(angle = 45, size = 3, nudge_x = 2, nudge_y = 0.06, family = "serif")+
  labs(x = 'Sample size (% of nodes from baseline activity profile)', y = 'Compliance', title = "Compliance between stable state of calibrated models (DP) \nand baseline activity profile")+
  ylim(0, 1)+
  scale_x_continuous(n.breaks = 10)

print(DP_compliance_plot)
```

### Stable state analysis of simulations using Z-score-based calibration datasets

```{r}
Z_data_frame = across_simulation_data_frame("Z", pipeline_folder_name)
Z_heatmap = across_simultation_heatmap(baseline_activity_file, "Z", pipeline_folder_name)
```

```{r}
consistent_nodes(Z_data_frame)
```

```{r}
Z_compliance = across_simulation_compliance(baseline_activity_file, "Z", pipeline_folder_name)

Z_compliance_plot = ggplot(Z_compliance, aes(x=size, y=Compliance, label = Compliance))+
  geom_point()+
  geom_hline(yintercept = random_model_compliance, color = "red", linetype = "dashed")+
  geom_text(aes(x = 100, y = random_model_compliance - 0.05, label = "Random model"), size = 3, family='serif', color = "red")+
  theme(text = element_text(size = 12, family = "serif"))+
  geom_text(angle = 45, size = 3, nudge_x = 2, nudge_y = 0.06, family = "serif")+
  labs(x = 'Sample size (% of nodes from baseline activity profile)', y = 'Compliance', title = "Compliance between stable state of calibrated models (Z-score) \nand baseline activity profile")+
  ylim(0, 1)+
  scale_x_continuous(n.breaks = 10)

print(Z_compliance_plot)
```

### Stable state analysis of model calibrated to near-minimal FVS

```{r}
FVS_data_frame = single_simulation_data_frame("FVS_simulation", pipeline_folder_name)

FVS_model_heatmap = single_simulation_heatmap("FVS_model_simulation", baseline_activity_file, "Stable state of model calibrated to near-minimal FVS")

FVS_compliance = single_simulation_compliance("FVS_simulation", baseline_activity_file, pipeline_folder_name)
```
